---
title: "Use of RAP package"
author: "Bart-Jan van Rossum"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: false
    number_sections: true
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Use of RAP package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
library(RAP)
```

# General introduction about package  

TO BE WRITTEN 

----

# Create TD object

The first step when modeling a field trial or performing a GxE analysis using 
the RAP package will always be creating an object of class TD. Such an object
is used internally by the package for performing all types of analysis.  

A TD object can be easily created from a data.frame by calling the function
`createTD`. This function will do a several things:  
* Rename columns to default column names used by the RAP package. For example, 
the column in the data containing variety/accession/genotype will allways be 
renamed to genotype. Original column names will be stored as an attribute to 
the final object of class TD.  
* Convert column type to the default column types. For example the column 
genotype will always be converted to a factor and rowCoord will always be 
converted to a numeric column.  
* Split the data by trial. The output TD object will be a list of data.frames
where each data.frame contains the data for a single trial. If there is only 
one trial or no column trial is defined the output will be a list with one
item only.  
* Add meta data to the seperate data.frames in the TD object. This meta data 
consists of location, date, longitude, latitude, plot design, width and length.
None of this is strictly neccessary for doing analysis but it is used for 
plotting field layouts, plotting trials on a map and naming plots. For a single 
trial can simply be added when creating the TD object using the function 
parameters in `createTD`. However if the data consists of multiple trials with
different meta data per trial it is more convenient to first create a TD object
without using the parameters for adding meta data. Doing so only a location will
be added based on the name of the trial. After creating the TD object the meta
data can be extracted using `getMeta`. This function outputs a data.frame with 
the meta data per trial. The content of this data.frame can then be modified 
and added back to the TD object using `setMeta`.  

After creating a TD object data for new trials can be added to it using 
`addTD`. This function works in exactly the same way as `createTD` except that
it adds data to an existing TD object instead of creating a new one. Still the
output needs to be assigned to a variable.  

As an example we will create a TD object using field data from an experiment
with wheat in Chili described by @Lado2013. This data will be used as an 
example throughout this vignette.  

The experiment was performed in two locations with 2 different drought regimes 
in two years for the first location and 1 trial for the second location. For the
sake of the example we will create an TD object for the first location first and
add the data for the second location later on. In practice this could also be 
done in one go.
```{r createTD}
data("wheatChl")
wheatTD <- createTD(data = wheatChl[wheatChl$trial != "C_SWS_12", ], 
                    genotype = "trt_id", repId = "rep", subBlock = "bl", 
                    rowId = "row", colId = "col", rowCoord = "row", 
                    colCoord = "col")
```
Note that both row and col are used twice. This is done since whenever a spatial
model is fitted both a factor column containing the spatial information and a
numeric column containing this information are needed. `rowId` and `colId` will 
be converted to a factor and `rowCoord` and `colCoord` will remain numerical 
column as they were in the input.  

The TD object we just created will be a list with 4 items, 1 for each trial 
(combination of location, drought regime and year) in the original data.frame. 
The meta data will be a data.frame with 4 rows.
```{r getMeta}
wheatMeta <- getMeta(TD = wheatTD)
head(wheatMeta)
```

Now we can modify the meta data and then add it back to the original TD object.
```{r setMeta}
wheatMeta$trLocation <- "Santa Rosa"
wheatMeta$trDate <- as.Date(rep(c("310811", "310812"), times = 2), "%d%m%y")
wheatMeta$trLat <- -36.32
wheatMeta$trLong <- -71.55
wheatMeta$trPlotWidth = 2
wheatMeta$trPlotLength = 1
wheatTD <- setMeta(TD = wheatTD, meta = wheatMeta)
```

To add the final trial to the TD object the function `addTD` can be used. Since
in this case we are only adding 1 new trial it makes sense to immediately add 
the meta data for this trial as well. 
```{r addTD}
wheatTD <- addTD(TD = wheatTD, data = wheatChl[wheatChl$trial == "C_SWS_12", ], 
                 genotype = "trt_id", repId = "rep", subBlock = "bl", 
                 rowId = "row", colId = "col", rowCoord = "row", 
                 colCoord = "col", trLocation = "Cauquenes", 
                 trDate = as.Date("070712", "%d%m%y"), trLat = -35.58,
                 trLong = -72.17, trPlotWidth = 2, trPlotLength = 1)
## Inspect the meta data after the extra trial was added.
getMeta(TD = wheatTD)
```

To get an idea of the content of the data and the field layout of the data we 
can summarize and plot the TD object. Multiple traits can be summarized at once 
but in order to save space we only look at GY (grain yield).
```{r TDsum}
summary(wheatTD, trial = "SR_FI_11", traits = "GY")
```

```{r layoutPlot}
plot(wheatTD, trials = "SR_FI_11")
```

The default plot creates plots for the layout of all trials in the TD object.
This can be restricted to selected trials using the `trials` parameter. The 
width and length of the plot will be derived from the trPlotWidth and 
trPlotLength in the meta data if these are available.  

A second type of plot will plot the trial locations on a map. This plot is made
based on trLat and trLong in the meta data. If not available the corresponding
location will not be plotted.
```{r mapPlot}
plot(wheatTD, plotType = "map")
```

----

# Modeling

After creating a TD object the next step will usually be fitting a model on
the trial data. This can be done using the function `STRunModel`. Depending on 
the design of the trial several different models can be fitted. The design can 
be given as a parameter in the function or can be included in the meta data of 
the TD object. In the first case the same model will be fitted for all trials, 
in the latter case different models can be fitted for different trials 
depending on the design.  

All models can be fitted using 3 different model fitting engines, 
SpATS [@RodAlv2017], lme4 [@Bates2015] and asreml. For models with rowcol or
res.rowcol design SpATS is the default 
engine, for the other models lme4. This can be overruled by the function 
parameter `engine`.

The exact models fitted are the following  
* ibd trait = genotype + *subBlock* + e  
* res.ibd trait = genotype + **repId** + **repId:subBlock** + e  
* rcbd trait = genotype + **repId** + e  
* rowcol trait = genotype + *rowId* + *colId* + e  
* res.rowcol trait = genotype + **repId** + *repId:rowId* + *repId:colId* + e  

In the models above fixed effects are indicated in *italics* whereas random 
effects are indicated in **bold**. genotype will be fitted as fixed or as 
random effect, or both, depending on the value of the parameter `what`. Extra
fixed effects may be fitted using the parameter `covariates`.  
In case SpATS is used as engine for fitting models an extra spatial term is 
always included in the model. The same is done when the engine is asreml and 
the function parameter `trySpatial` is set to true.

Using the TD object wheatTD we created in the previous section we can now fit 
a model on the data. 
```{r fitSp, message=FALSE}
modWheat <- STRunModel(TD = wheatTD, trials = "SR_FI_11", traits = "GY",
                       design = "res.rowcol")
```
Note that by not supplying the `what` argument to the function two models will
be fitted, one with genotype as a fixed effect and one with genotype as a 
random effect. Because of this both BLUEs and BLUPs can be calculated from 
the output and this will be done when summarizing the results. If genotype 
would have been fitted as fixed effect only, the summary would only have 
contained BLUEs. When fitting genotype as random effect only the summary would 
only have contained the BLUPs.
```{r fitSum, message=FALSE}
## Set nBest to 5 to decrease size of output.
summary(modWheat, nBest = 5)
```



----

# Extract 

Explanation on what can be extracted from a model and how

----

# References


