---
title: "Modeling field trials using RAP"
author: "Bart-Jan van Rossum"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: false
    number_sections: true
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Modeling field trials using RAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.dim = c(6, 4)
)
library(RAP)
```

# The RAP Package  

The RAP package is developed as an easy-to-use package for analyzing data of 
plant breeding experiments with many options for plotting and reporting 
the results of the analyses.  

The package has three main components:

* Single trial analysis
* Genotype by Environment (GxE) analysis
* QTL analysis

This vignette deals with the modeling part of the package and describes in 
detail how to prepare data for analysis, perform analyses using different
modeling engines and extract the results from the models.

----

# Data preparation

The first step when modeling field trial data or performing a GxE analysis with 
the RAP package is creating an object of class `TD` (Trial Data). This object is used by the RAP package for performing most analyses.  

## Creating a `TD` object

A `TD` object can be created from a `data.frame` with the function `createTD`. 
This function does a number of things:  

* Rename columns to default column names used by the functions in the RAP package. 
For example, the column in the data containing variety/accession/genotype is renamed to 
"genotype". Original column names are stored as an attribute of the `TD` object.
* Convert column types to the default column types. For example, the column 
"genotype" is converted to a factor and "rowCoord" to a numeric column.
* Split the data into separate data.frames by trial. A `TD` object is a
`list` of `data.frames` where each `data.frame` contains the data for a single
trial. If there is only one trial or no column trial is defined, the output will 
be a `list` with only one item.
* Add meta data to the separate `data.frames` in the `TD` object. The meta data 
consists of location, date of the experiment, longitude, latitude, trial design,
plot width and plot length. None of these is strictly neccessary for any 
analysis but the meta data is used for plotting field layouts, plotting trials 
on a map and naming plots. Meta data for a single trial can simply be added when creating the `TD` object using the parameters in `createTD`. However if the data consists of multiple trials with different meta data per trial it is more 
convenient to first create a `TD` object without adding meta data. This will 
still add a location based on the name of the trial. After creating the `TD` 
object, the meta data can then be extracted using `getMeta`. This function 
creates a `data.frame` with the meta data per trial. The content of this 
`data.frame` can then be modified and added back to the `TD` object using 
`setMeta`.  

After creating a `TD` object, data for new trials can be added to it using 
`addTD`. This function works in exactly the same way as `createTD` except that
it adds data to an existing `TD` object instead of creating a new one.  
Dropping one or more trials from a `TD` object can be done using the function
`dropTD` specifying the trials to be dropped in the parameter `rmTrials`.

## Example

Field data from an experiment with wheat in Chile described in detail by 
@Lado2013 will be used as an example throughout this vignette. The experiment 
was performed in two locations in Chile, Santa Rosa and Cauquenes, with two 
different drought regimes in 2011 and 2012 for the Santa Rosa and one trial 
in 2012 for Cauquenes. For 384 genotypes four traits were measured in 2011, but in 
2012 only grain yield (GY) was measured and the examples will mainly focus on 
this trait.  

For the example first a `TD` object is created for the first location and 
the data for the second location is then added later on. In practice all this
could be done in one go.
```{r createTD}
## Create a TD object containing the data from Santa Rosa.
data("wheatChl")
wheatTD <- createTD(data = wheatChl[wheatChl$trial != "C_SWS_12", ], 
                    genotype = "trt", repId = "rep", subBlock = "bl", 
                    rowCoord = "row", colCoord = "col")
```
The `TD` object just created is a `list` with four items, one for each trial 
(combination of location, drought regime and year) in the original `data.frame`.
The column "trt" in the original data is renamed to "genotype" and converted to 
a factor. The columns "rep" and "bl" are renamed and converted likewise.
The columns "row" and "col" are renamed to "rowCoord" and "colCoord" 
respectively. Simultaneously two columns "rowId" and "colId" are
created containing the same information converted to a factor. This seemingly
duplicate information is needed for spatial analysis. It is possible to define
different columns as "rowId" and "colId" than the ones used as "rowCoord" and
"colCoord". The information about which columns have been renamed when creating a `TD` object is stored as an attribute of each individual `data.frame` in the object.

## Meta data

The meta data will be a `data.frame` with four rows, one for each item in `wheatTD`.
```{r getMeta}
## Extract meta data from the TD object. 
(wheatMeta <- getMeta(TD = wheatTD))
```

After extracting the meta data, it can be modified and then added back to the original `TD` object.
```{r setMeta}
## Fill in meta data and add back to the TD object.
wheatMeta$trLocation <- "Santa Rosa"
wheatMeta$trDate <- as.Date(rep(c("310811", "310812"), times = 2), "%d%m%y")
wheatMeta$trLat <- -36.32
wheatMeta$trLong <- -71.55
wheatMeta$trPlWidth = 2
wheatMeta$trPlLength = 1
wheatTD <- setMeta(TD = wheatTD, meta = wheatMeta)
```

## Add extra data to a `TD` object

To add the data for the final trial to the `TD` object the function `addTD` can
be used. Since now only one new trial is added, it makes sense to immediately add the 
meta data for this trial as well, using the appropriate parameters in `addTD`. 
```{r addTD, R.options=list(width=90)}
## Add the data for Cauquenes to the TD object.
wheatTD <- addTD(TD = wheatTD, data = wheatChl[wheatChl$trial == "C_SWS_12", ], 
                 genotype = "trt", repId = "rep", subBlock = "bl", 
                 rowCoord = "row", colCoord = "col", trLocation = "Cauquenes", 
                 trDate = as.Date("070912", "%d%m%y"), trLat = -35.58,
                 trLong = -72.17, trPlWidth = 2, trPlLength = 1)
## Inspect the meta data after the extra trial was added.
getMeta(TD = wheatTD)
```

## Summarizing a `TD` object

The `summary` function can be used to get an idea of the content of the data in the `TD` object. Multiple traits may be summarized at once but for clarity here a summary is only made for GY.
```{r TDsum}
## Create a summary for grain yield in SR_FI_11.
summary(wheatTD, trial = "SR_FI_11", traits = "GY")
```

Using the default options, nine summary statistics are printed, but many more are 
available. These can be accessed using the parameter `what` in the `summary`
function. For a full list of available statistics, use `help(summary.TD)`. It is 
also possible to output all statistics using `what = "all"`.    

It is possible to summarize the data in a TD object for different groups. This
can be done using the parameter `groupBy`. It will display three main summary
statistics per group. Again, more statistics can be displayed using the 
parameter `what`.
```{r TDsumGroup}
## Create a summary for the two replicates in SR_FI_11.
summary(wheatTD, trial = "SR_FI_11", traits = "GY", groupBy = "repId")
```

## Plotting a `TD` object

Several plots can be made to further investigate the contents of a `TD` object.    

The default plot creates plots for the layout of all trials in the `TD` object.
This can be restricted to selected trials using the `trials` parameter. The 
width and length of the plot are derived from "trPlWidth" and "trPlLength" in 
the meta data if these are available.  
```{r layoutPlot}
plot(wheatTD, trials = "SR_FI_11")
```

This plot can be extended by highlighting interesting genotypes in the layout.
```{r layoutPlotHL}
## Plot the layout for SR_FI_11 with genotypes G278 and G279 highlighted.
plot(wheatTD, trials = "SR_FI_11", highlight = c("G278", "G279"))
```

It is also possible to color the subBlocks within the field.
```{r layoutPlotSB, fig.dim = c(6, 5)}
## Plot the layout for SR_FI_11, color subBlocks.
plot(wheatTD, trials = "SR_FI_11", colorSubBlock = TRUE)
```

Highlighting genotypes and coloring subBlocks cannot be done simultaneously. If 
both options are specified, only highlighting is done.    

Finally, it is possible to add the names of the genotypes to the layout. 
```{r layoutPlotSG, fig.dim = c(6, 5)}
## Plot the layout for SR_FI_11, color subBlocks.
plot(wheatTD, trials = "SR_FI_11", showGeno = TRUE)
```

A second type of plot displays the trial locations on a map. This plot is made 
based on trLat and trLong in the meta data. If latitude or longitude is not available 
for a certain location, then this location is not plotted. If the locations are 
very close together, the resulting map can become quite small. The parameters 
`minLatRange` and `minLongRange` can be used to extend the minimum range of latitude
and longitude respectively to address this issue.
```{r mapPlot}
## Plot the locations of the trials on a map.
plot(wheatTD, plotType = "map")
```

Boxplots can be made to get an idea of the contents of the data in the `TD` object.
By default a box is plotted per trial in the data for the specified traits. Boxplots for multiple traits can be made at once.
```{r boxPlot}
## Create a boxplot for grain yield.
plot(wheatTD, plotType = "box", traits = "GY")
```

The trials in the boxplot can be grouped using the parameter `groupBy`. Colors
can be applied to groups within trials using the parameter `colorBy`. The boxes 
for the (groups of) trials can be ordered using `orderBy`. Boxes can be ordered by external vector or by the group mean.
```{r boxPlotGR}
## Create a boxplot for grain yield with boxes grouped by year and repIds within
## years colored.
plot(wheatTD, plotType = "box", traits = "GY", groupBy = "year", 
     colorBy = "repId", orderBy = "descending")
```

The last plot that can be made is a plot of the correlations between the fields for
a specified trait. The order of the plotted fields is determined by clustering them
and plotting closely related field close to each other.
```{r corPlot}
## Create a correlation plot for grain yield.
plot(wheatTD, plotType = "cor", traits = "GY")
```

----

# Modeling

After creating a `TD` object, a model can be fitted on the trial data. This is done 
using the function `fitTD`. Depending on the design of the trial several 
different models can be fitted. The design can be given as a parameter in the 
function or included in the meta data of the `TD` object. In the former case the 
same model will be fitted for all trials, in the latter different models can be 
fitted for different trials depending on the design. If both are available the 
function parameter will always be leading. 

The output of `fitTD` is an object of class `SSA` (Single Site Analysis), a
`list` of fitted models with one item for each trial the model was fitted for. 

## Model types

`fitTD` uses three different engines for fitting the models, namely
SpATS [@RodAlv2017], lme4 [@Bates2015] and asreml [@Gilmour2017]. For models 
with row column or resolvable row column design, SpATS is the default engine, 
for the other models lme4. This can always be overruled by specifying the function 
parameter `engine`.

Models can be fitted for five different trial designs. These are listed in the 
following table with their respective model specifications.  

design | code | model fitted |
-------------------------- | -------- | ----------------------------------------- |
incomplete block design | ibd | trait = genotype + *subBlock* + $\epsilon$ |
resovable incomplete block design | res.ibd | trait = genotype + **repId** + **repId:subBlock** + $\epsilon$ |
randomized complete block design | rcbd | trait = genotype + **repId** + $\epsilon$ | 
row column design | rowcol | trait = genotype + *rowId* + *colId* + $\epsilon$ |
resolvable row column design | res.rowcol | trait = genotype + **repId** + *repId:rowId* + *repId:colId* + $\epsilon$ |

In the models above, fixed effects are indicated in *italics* whereas random 
effects are indicated in **bold**. genotype can be fitted as fixed or as 
random effect depending on the value of the parameter `what`. Extra
fixed effects may be fitted using the parameter `covariates`.  
If SpATS is used as modeling engine, an extra spatial term is always 
included in the model. The same is done when the engine is asreml and 
the function parameter `trySpatial` is set to `TRUE`.

Using the `TD` object wheatTD from the previous section, a model for the trial 
SR_FI_11 and trait GY can now be fitted on the data as follows. For the design
of the trial a resolvable row column design is assumed. Since `engine` is not 
supplied as a parameter SpATS is used for fitting the model.
```{r fitSp, message=FALSE}
## Fit a single trial model.
modWheatSp <- fitTD(TD = wheatTD, trials = "SR_FI_11", traits = "GY", design = "res.rowcol")
```
Note that by not supplying the `what` argument to the function, two models are
fitted, one with genotype as a fixed effect and one with genotype as a 
random effect. The results of both these models are stored in the `SSA` object
`modWheatSp`. This is very useful for extracting different results from the 
model later on. A trade-off is that fitting two models takes more time than 
fitting only one so when fitting models on large datasets it is sensible to explicitly define `what` if only a subset of the results are needed as output.
```{r fitSpSm, message=FALSE}
## Fit a single trial model with genotype as random effect.
modWheatSp2 <- fitTD(TD = wheatTD, trials = "SR_FI_11", traits = "GY", 
                     what = "random", design = "res.rowcol")
```

## Spatial models

When using SpATS as a modeling engine for fitting a model, an extra spatial 
component is always included in the model. This spatial component is composed
using the `PSANOVA` function in the SpATS package which uses 2-dimensional 
smoothing with P-splines as described in @Lee2013 and in @RodAlv2017. See `help(PSANOVA, SpATS)`
for a detailed description. The parameters `nseg` and `nest.div` of `PSANOVA` 
can be modified using the `control` parameter in `fitTD`.  
Fitting the model in the previous section specifying the number of segments
for both rows and columns as 20 works as follows
```{r fitSpCtr, message=FALSE}
## Fit a spatial single trial model using SpATS. 
## Manually specify the number of segments for rows and columns.
modWheatSp3 <- fitTD(TD = wheatTD, trials = "SR_FI_11", traits = "GY", 
                     design = "res.rowcol", control = list(nSeg = c(20, 20)))
```

Alternatively, spatial models can be fitted using asreml as modeling engine 
and setting the parameter `trySpatial = TRUE`. In this case seven models are fitted 
and the best model, based on a goodness-of-fit criterion, either AIC or BIC, is 
chosen. For a full specification of the models fitted see `help(fitTD)`. 
The criterion to be used can be specified using the `control` parameter in `fitTD`.  
Fitting a model similar to the one above using asreml with BIC as 
goodness-of-fit criterion works as follows
```{r fitAs, message=FALSE, results='hide'}
if (requireNamespace("asreml", quietly = TRUE)) {
  ## Fit a spatial single trial model using asreml.
  modWheatAs <- fitTD(TD = wheatTD, trials = "SR_FI_11", traits = "GY", design = "res.rowcol", trySpatial = TRUE, engine = "asreml", control = list(criterion = "BIC"))
}
```
The fitted models and the best model are stored in the output together with a summary
table with details on the fitted models.
```{r spatCh}
if (exists("modWheatAs")) {
  ## Overview of fitted models
  print(modWheatAs$SR_FI_11$sumTab$GY, digits = 2)
}  
```
`r if (exists("modWheatAs")) {"As the overview shows the best model, the model with the lowest BIC, is AR1(x)AR1."}`

## Model output

Since genotype has been fitted both as fixed and as random factor in 
`modWheatSp` it is possible to calculate both the Best Linear Unbiased Estimators 
(BLUEs) and the Best Linear Unbiased Predictors (BLUPs). Therefore both are printed 
in the summary of the model together with their respective standard errors.
```{r fitSum, message=FALSE}
## Set nBest to 5 to decrease size of output.
summary(modWheatSp, nBest = 5)
```

## Model plots

Two types of plots can be made for fitted models. The first is a series
of four plots, a histogram of the residuals, normal quantiles of the residuals, 
a scatter plot of residuals against fitted values and a scatter plot of absolute
value of residuals against fitted values. These plots can be made by calling 
`plot` on the `SSA` object. Plots can be made for multiple trials and multiple traits 
simultaneously, either for the model with genotype as fixed effect or for the model 
with genotype as random effect. By default plots are made for all trials and all traits, 
but this can be restricted using the parameters `trials` and `traits`. If only one model is fitted, i.e. only the model with genotype as fixed effect or only the model with genotype as random effect, the results of this model will be plotted. In case both models were fitted, as a default the results will be plotted for the model with genotype fixed. This can be changed using the parameter `what`.
```{r basePlot}
## Base plots for the model with genotype fitted as random effect.
plot(modWheatSp, what = "random")
```

The second type of plot consists of five plots, spatial plots of the raw data, 
fitted values, residuals and either BLUEs or BLUPs, and a histogram of the BLUEs 
or BLUPs. If SpATS was used for modeling an extra plot with the fitted spatial trend is included. The call for creating these plots differs from the base plots only by an 
extra parameter `plotType = "spatial"`. Note that spatial plots can only be made if spatial information, i.e. `rowCoord` and `colCoord`, is available in the `TD` object.
```{r spatPlot}
## Spatial plot for the model with genotype fitted as fixed effect.
plot(modWheatSp, plotType = "spatial")
```

## Model reports

As with many objects in the RAP package there is a `report` function available 
for `SSA` objects. This `report` function creates a pdf report describing 
the main results of the fitted model. Also the tex file and figures used for generating
the pdf report are saved. By editing the tex file it is possible to modify the report to ones needs, creating high flexibility.  

When no outfile is specified, a report will be created with a default name, i.e. "modelReport.pdf", in the current working directory. The parameter `outfile` can be used to change the location. This value of this parameter should be a valid location and name for a **pdf** file, i.e. including the postfix ".pdf". Non-existing directories are created by the `report` function.  

The report contains general information on the model fitted, a summary of the results, 
the plots described in the previous section, a `list` of best (highest BLUEs or BLUPs) genotypes and a scatter plot of all genotypes and their BLUEs or BLUPs. If, for the modeled trait, a low value means the genotype is performing well, then the selection on the best genotypes can be based on the lowest BLUEs or BLUPs by setting the parameter `descending` to `TRUE` in the report function.
```{r modRep, eval=FALSE}
## Create a report in the current working directory
report(modWheatSp)
## Create a report for the model with genotype fitted as random.
report(modWheatSp, outfile = "./myReports/wheatReport.pdf", what = "random")
```
Reporting for an `SSA` object is always done for one trait in one trial for one fitted model. Parameters `trait`, `trial` and `what` can be used for selecting the model for
which the report should be made.

----

# Extracting model results

After fitting a model, various results can be extracted or calculated from the 
fitted model using the function `extract`. This can be anything from a single 
result for one trait and one trial to a `list` of different results for all models 
in an `SSA` object. The results that can be extracted depend on the type of 
model fitted and sometimes on the modeling engine as well. For example, BLUEs 
can only extracted if genotype was fitted as fixed effect whereas BLUPs and heritabilities can only be extracted and calculated if genotype was fitted as random effect.  

All results that can be extracted are shown in the table below. The first column
shows what model needs to be fitted in order to be able to extract the result.
Here F stands for genotype as fixed effect and R for genotype as random effect. In the 
second column is the result. This is also the value to be used for the parameter
`what` in `extract` needed to extract the corresponding result. The last 
column gives a short description of the result that will be extracted and, where
needed, also states for which modeling engines it can be extracted.

model | result | description |
----- | ------ | ------------------------------------------------------------- |
F | BLUEs | Best Linear Unbiased Estimators |
F | seBLUES | standard errors of the BLUEs |
R | BLUPs | Best Linear Unbiased Predictors |
R | seBLUPs | standard errors of the BLUPs |
F | ue | unit errors - only for `lme4` and `asreml` |
R | heritability | broad sense heritability |
F | varCompF | variance components for the model with genotype as fixed component |
R | varCompR | variance components for the model with genotype as random component |
R | varGen | genetic variance component |
R | varErr | residual variance component - only for `lme4` and `asreml` |
R | varSpat | spatial variance components - only for `SpATS` |
F | fitted | fitted values for the model with genotype as fixed component |
F | resid | residuals for the model with genotype as fixed component |
F | stdRes | standardized residuals for the model with genotype as fixed component |
R | rMeans | fitted values for the model with genotype as random component |
R | ranEf | random genetic effects |
R | residR | residuals for the model with genotype as random component |
R | stdResR | standardized residuals for the model with genotype as random component |
F | wald | results of the wald test - only for `lme4` and `asreml` |
F | CV | Coefficient of Variation - only for `lme4` and `asreml` |
F | rDf | residual degrees of freedom for the model with genotype as fixed component |
R | rDfR | residual degrees of freedom for the model with genotype as random component |
R | effDim | effective dimensions - only for `SpATS` |
R | ratEffDim | ratios of the effective dimensions - only for `SpATS` |
F | sed | standard error of difference - only for `asreml` |
F | lsd | least significant difference - only for `asreml` |

Using `what = "all"` in the function call, extracts all results possible for the
fitted model. This is also the default.  

Below are some examples of extracting results from a fitted model. Recall that `modWheatSp` contains two fitted models, one with genotype as fixed effect and one with genotype as random effect.
```{r extBLUEs}
## Extract BLUEs
BLUEsWheat <- extract(SSA = modWheatSp, what = "BLUEs")
## Extract BLUEs and BLUPs
predWheat <- extract(SSA = modWheatSp, what = c("BLUEs", "BLUPs"))
```
Both `BLUEsWheat` and `predWheat` are a `list` with one item, the trial used for
modeling. In general, when extracting results, there will be an item in the `list` for every trial for which results were extracted. These items are a `list` themselves with an item for every statistic that has been extracted, so one item for `BLUEsWheat`, a `data.frame` containing the BLUEs, and two for `predWheat`, the `data.frames` containing BLUEs and BLUPs respectively. 

## Adding extra variables

The `data.frame` BLUEs in either of the `lists` consists of only two columns, 
genotype and GY. If the model would have been fitted for multiple traits all 
these traits would become columns in the `data.frame`. It might be useful to add
extra columns from the data used to fit the model to the output. This can be 
achieved using the paramater `keep` in `extract`. To include the trial in the 
output, useful for easily combining several `data.frames` with BLUEs and using
them for a GxE analysis, use the following command
```{r extBLUEsKeep}
## Extract BLUEs from the fitted model.
BLUEsWheat2 <- extract(SSA = modWheatSp, what = "BLUEs", keep = "trial")
head(BLUEsWheat2[["SR_FI_11"]]$BLUEs)
```
Not every column from the original `TD` object can be included in the extracted 
data in this way. Only columns for which the values are uniform per genotype can 
be included. For example, the column `repId` containing replicates, that has several different values for a single genotype, cannot be included. When trying to do so it will 
be dropped with a warning.  
It is however possible to include `repId` when extracting fitted values, since 
for each observation in the original data a fitted value is computed. 
```{r extFit}
## Extract fitted values from the model.
fitVals <- extract(SSA = modWheatSp, what = "fitted", keep = c("trial", "repId"))
head(fitVals[["SR_FI_11"]]$fitted)
```

## Prepare data for GxE analysis

To use the BLUEs or BLUPs from the fitted model in a GxE analysis they have to be 
converted again into a `TD` object. For this the function `SSAtoTD` can be used. It
creates a `TD` object from a fitted model outputting one or more of the following:
BLUEs, standard errors of BLUEs, BLUPs and standard errors of BLUPs. Optionally, a column `wt` with weights (calculated as $(1 / seBLUEs)^2$) can be added as well. In the same way as described in the previous section extra columns can be added to the output using the parameter `keep`. 
```{r SSAtoTD, message=FALSE}
## Fit a model for all trials with genotype as fixed factor.
modWheatSpTot <- fitTD(TD = wheatTD, traits = "GY", what = "fixed", 
                       design = "res.rowcol")
## Create a TD object containing BLUEs and standard errors of BLUEs.
TDGxE <- SSAtoTD(SSA = modWheatSpTot, what = c("BLUEs", "seBLUEs"))
## Add weights to the output.
TDGxE2 <- SSAtoTD(SSA = modWheatSpTot, what = c("BLUEs", "seBLUEs"), addWt = TRUE)
```

See the vignette [Genotype by Environment analysis using RAP](RAPGxE.html) for an overview of how to use the input just created for performing GxE analysis.

----

## References


